Debugging by \textit{lastChange}

### 2010-10-28 papers/pldi2011/sigplanconf-template.tex
smaller edits on section 3 - 5

\section{Introductory example}

@@

\begin{figure}[htp]
\includegraphics{5-example-points.jpg}
\caption{The examined points before locating the defect. The arrow represents the logical forward progress of the program. Three actual executions are superimposed on this arrow. All three stop at the reproduction point indicated by the circle 1. After the first execution, the developer asks for lastChange as described in \secref{introExample}, yielding information  indicated by circle 2. After the second execution, another lastChange query causes a third execution, yielding information indicated by circle 3.}
\label{fig:example-points}
\end{figure}

@@

\subsection{\textit{lastChange} on object propery}
To simplify the algorithm explanation and defer technical details, we
define two basic operations and later we explain the details of these
two operations. The first opertion is \texttt{objectId()}: given a
JavaScript object it returns an integer as its identifier. This
identifier is unique to the object during one execution. The second
operation is \texttt{setPropertyChangeHook()}: given a function and a
string, the function is called whenever a property changes and its
name matches the string For example, if the property might be
\texttt{foo} so changes to \texttt{bar.foo} or \texttt{baz.foo} would
call the function.  The callback function receives a reference to the
owner of \texttt{foo}. In the next section, we describe how we use
a slightly different version of this operation which is restricted 
to only created objects in one location instead of all objects.

@@

To see how these functions work, suppose the developer asks for the
last change of \texttt{bar.foo} at the reproduction point in a
program. The debugger calls \texttt{setPropertyChangeHook()} with
\texttt{foo} as the property name and re-executes the
program. Whenever \texttt{foo} changes and the callback function is to
be called, debugger first calls \texttt{objectId()} on the
\texttt{foo} owner object. Then it stores this owner id, the stack
frame locations, and other state values in scope at the call point.
Whenever the execution reaches the reproduction point the debugger
looks at the history of \texttt{foo} changes and finds the last
\texttt{foo} change with the same object id as \texttt{bar} id at the
reproduction point. Figure ~\ref{fig:foo-changes1} shows the list of
property \texttt{foo} change events in a hypothetical
execution. \texttt{bar} id at the reproduction point is 1010, so the
last change of \texttt{bar.foo} is the fourth column.  By using an
object id instead of an object reference we allow the garbage
collector to reclaim the space for dead objects just at would in the
absence of the debugger.

\subsection{\textit{lastChange} on variable} 
In JavaScript, every frame has a scope chain and every available
variable in the frame comes from one of the scopes in the frame's
scope chain. Once developer asks for the last change of variable
\texttt{foo} at the reproduction point, debugger first determines the
variable's scope as follows: it iterates over the scopes in the scope
chain and the first scope which has a variable with the same name is
the variable's scope. There are five different scope types: local,
global, \texttt{with}, closure and \texttt{catch}. We explain these
cases in two groups.

\subsubsection{global and \texttt{with} scopes}
Global scope is the the most outer scope in the scope chain and it is
also referred to as the global object (the \texttt{window}
object in Web pages). This scope is a regular JavaScript object and therefore every
global variable is a property of global object. Similarly,
 \texttt{with} scopes are also regular objects. A \texttt{with}
scope is created by a \texttt{with()} block with an object as the
parameter. Every property of this parameter object is available inside the
block as a variable. Debugger treats the case where variable's scope
is global or \texttt{with}, like \textit{lastChange} on an object
property.

@@

If developer asks for the last change of variable \texttt{foo} at the 
reproduction point in a program, debugger calls 
\texttt{setVariableChangeH-} \texttt{ook()} with the variable's defining
block and name as parameters and re-executes the
program. Whenever \texttt{foo} changes and the callback function is to
be called, debugger first calls \texttt{scopeId()} on the
variable's scope. Then it stores this scope id, the stack
frame locations, and other state values in scope at the call point.
Whenever the execution reaches the reproduction point the debugger
looks at the history of \texttt{foo} changes and finds the last
\texttt{foo} change with the same scope id as the variable's scope id
at the reproduction point (Figure ~\ref{fig:foo-changes2}).

@@

\subsection{\textit{lastChange} on \textit{lastChange}}
The algorithm we explained so far needs an enhancement to support
longer sequence of \textit{lastChange}s. We explain first the issue
and then the solution. Consider the following points:

\begin{center}
\textit{
 point A : the reproduction point \\
 point B : lastChange(A, bar.x) \\
 point C : lastChange(B, baz.y) 
 }
 \end{center}
According to the algorithm explained in the previous section none of
the points B and C can be identified before reaching point
A. Therefore, all \texttt{x} change events for point B and all \texttt{y}
change events for point C are stored. Once the execution reaches point A,
and the exact \texttt{bar} id reveals, point B can be identified among
the change events in the history of \texttt{x} changes. The problem
arises in recognizing point C. To recognize point C, we need
\texttt{baz} id at point B, but B is a past program state, and
\texttt{baz} id has not been collected at this point.

To address this issue, we use Algorithm \ref{dependency-analysis}, to
create the list of additional data should be collected at a change
event. This algorithm does does a dependency analysis between all
\textit{lastChange} queries and computes the list additional data
required to be collected at each change event. For example in the
mentioned case, \texttt{baz} id-if it is available- should be
collected at all points \texttt{x} changes. Having this additional
data, locating point C, using \texttt{baz} id at point B is
straightforward (Figure ~\ref{fig:lastchange-lastchange}).

@@

%---------------------------------------------------------------------------------------------------
\section{JavaScript implementation}
To verify the \textit{lastChange} algorithm we
implemented it in an extension to the Firebug
JavaScript debugger\cite{Firebug}. %\cite{firebug-version-1.6}. 
Firebug itself is an extension of the Firefox browser\cite{Firefox}. %\cite{firefox-version-3.6}
The Firefox JavaScript engine provides a JavaScript debugging interface \cite{JSD} and
\textit{Querypointer} is developed over this interface.

@@

For our purposes, the \texttt{watch()} function only covers the case
of global object properties. At the beginning of execution, no object
excepting the global object is available. Therefore we defined and
implemented \texttt{setPropertyChangeHook()} to work in the rest of
the cases.

The Firefox JavaScript engine does not current support
\texttt{setPro-}
\texttt{pertyChangeHook()} (see however []).  For %\cite{bugzilla}
the our \textit{lastChange} prototype we created a version of
\texttt{setPropertyChangeHook()} using only features of the current
engine.  The basic strategy is to get a reference to the object just
after its creation, then use \texttt{watch()} function to monitor
property changes in the object.  Setting a flag into the Firefox
JavaScript engine we can get file URL and line number for each object
creation (e.g., myFile.js, line 24). We set a hook on this line
and analyze the source code to determine which object was created.

Here, we implicitly assumed that the object will be created at 
the same location in the next execution. So, what does happens if it
is not true? Well, once the execution reaches the reproduction point, it
reveals that the object has been created in different location. This time,
debugger re-executes the program considering both locations as possible 
object creation locations.

\subsection{Tracking object creation}
The only data we have is the object creation location including the
file url and the line number and the goal is to get a reference to
this object. Although in the most regular cases we have only one
statement and one object creation in a line, there are cases that more
than one object might be created in a line. There is no way to
recognize the interesting object among these new objects. So instead
of one object, we monitor all new objects created in the line.

@@

There is another case for new objects and when they are passed as
arguments to other functions Figure~\ref{fig:objectCreation}(c). In these
cases instead of step-over we use step-in and we get the corresponding
argument as a new object.

Although this approach is successful in many ordinary cases, there are some 
cases that this approach doesn't work. For example in a case
where the new object is assigned to \texttt{a[++i]} where i is a variable
(Figure ~\ref{fig:objectCreation}(d)), \texttt{a[++i]} is kept in the list
of assignee statements. Obviously, evaluating this statement doesn't return
a reference to the new object.

\subsection{\texttt{scopeId()} operation}
Debugger sets a hook at the beginning of all code blocks need
a scope id whenever the block is loaded. The scope id is kept as a variable with name 
\texttt{\_scopeId} in the scope. Whenever the hook hits which means a new scope
is created, \texttt{\_scopeId} is set by calling JavaScript's dynamic compilation 
function \texttt{eval()}. For example, executing \texttt{eval( var \_scopeId = 10)} creates a
variable with name \texttt{\_scopeId} and value \texttt{10} in the
scope of the \texttt{eval()} call, which is our interesting scope. \texttt{scopeId()}
operation returns the value of \texttt{\_scopeId} in the scope.

\subsection{\texttt{setVariableChangeHook()} operation}
Variables defined in local, closure and catch scopes are only change in
their scope; that includes the defining scope and scopes nested inside of them.
For example in Figure ~\ref{fig:js-closure}, variable \texttt{var1} in line 11 can only be
changed in lines 10 to 23. Therefore, after locating the function in
which the variable is defined, it is enough to parse the code inside
the function block and set a hook on all lines where the variable is
assigned a new value. 
We also set a hook at the first line of the function which is
corresponding to the line variable is defined. In Figure
~\ref{fig:js-closure}, if the execution is paused at line 17 and the
last change of \texttt{var1} is queried, two hooks on lines 16 and 17
will be set, but if the execution is paused at line 20 , four hooks on
lines 11, 12, 14, 20 will be set. These two cases are different due
the fact that \texttt{var1} in \texttt{firstChild} is a local variable
but in \texttt{secondChild} is a closure variable.

@@

\section{The effect of Non-determinism}  %flying over non-determinism
\subsection{Reproducible Non-deterministic Execution}

@@

Since \textit{lastChange} does not rely on determinism, some bugs that
confound breakpoint debuggers will be easier to find with
\textit{lastChange}.  Consider the following example.
Figure ~\ref{fig:counter-example} shows a function with processes an
array. The array contains numbers except one itme which is
\texttt{undefined} and it causes a bug in line 30. There is a call to
\texttt{randomPermutation} function in line 11 which randomly
permutes the array item. So at every execution the
\texttt{undefined} item will be in a new place. Calling
\textit{lastChange} on \texttt{x} at the place bug happens, gives a
point which shows line 14. Although this point exists at every
re-execution but it can not be identified by a conditional breakpoint.

@@

We have presented the \textit{lastChange} algorithm and describe our
prototype implementation. Now we want to convince you that the
algorithm has practical value. Ultimately, practical value can only be
proven by developers in the field using the technique. This paper is
one step to convince implementers to put \textit{lastChange} in the
field.  Thus our goal here is a persuasive argument.

@@

\subsection{developers need an operation like \protect\textit{lastChange} }

We have shown how \textit{lastChange} identifies the prior point in
program execution where a program state value changes. Is it something
the developer needs to do?  Since ultimately programs are just
transformation of state values, debugging is ultimately backtracking
to find defects in program state change.  When a developer halts a
program on a breakpoint they have three kinds of information: a model
of the program in their head, a call stack showing what part of the
program is halted, and the state at that point (a combination of the
debugger's view of the program internal state and the output of the
program up to this point of execution). If the model does not match
the call stack or if no value in the viewable program state aligns
with the developers model for this point in execution, the developer
will seek another view by changing the breakpoint or the input
data. If a value appears incorrect, they may have a flash of insight
and know the defect. Otherwise they need to figure out what operation
causes the incorrect value. Thus \textit{lastChange} addresses a key
part of the debugging process.

\subsection{developers can learn to use \protect\texttt{lastChange}}

@@

\subsection{practical implementations are feasible}

@@

\subsection{in most cases \protect\texttt{lastChange} will be much faster than current alternatives}

@@

We could also compare to solutions based on logging or tracing. Manual
logging has very high overhead: the developer must add code, debug
that added code, then analyze the log. (To be fair, the log can become
a permanent debugging aid.) Automatic logging as we discussion in
\serref{relatedWork} causes about one or two orders of magnitude slow down.

\subsection{the worst cases are not more common or more painful than alternatives}

@@

\section{Related Work}

@@

\section{Conclusions and Future Work}

>>>>

\section{Introductory example}
\label{sec:introExample}

@@

\begin{figure}[htp]
\includegraphics{5-example-points.jpg}
\caption{The examined points before locating the defect. The arrow represents the logical forward progress of the program. Three actual executions are superimposed on this arrow. All three stop at the reproduction point indicated by the circle 1. After the first execution, the developer asks for lastChange as described in \ref{sec:introExample}, yielding information  indicated by circle 2. After the second execution, another lastChange query causes a third execution, yielding information indicated by circle 3.}
\label{fig:example-points}
\end{figure}

@@

\subsection{\textit{lastChange} on object property}
To simplify the algorithm explanation and defer technical details, we
define two basic operations and later we explain the details of these
two operations. The first operation is \texttt{objectId()}: given a
JavaScript object it returns an integer as its identifier. This
identifier is unique to the object during one execution.  By using an
object id instead of an object reference we allow the garbage
collector to reclaim the space for dead objects just at would in the
absence of the debugger. The second
operation is \texttt{setPropertyChangeHook()}: given a function and a
string, the function is called whenever a property changes and its
name matches the string For example, if the property might be
\texttt{foo} so changes to \texttt{bar.foo} or \texttt{baz.foo} would
call the function.  The callback function receives a reference to the
owner of \texttt{foo}. In the next section, we describe how we use
a slightly different version of this operation which is restricted 
to only created objects in one location instead of all objects.

@@

To see how these functions work, suppose the developer asks for the
last change of \texttt{bar.foo} at the reproduction point in a
program. The debugger calls \texttt{setPropertyChangeHook()} with
\texttt{foo} as the property name and re-executes the
program. Whenever \texttt{foo} changes and the callback function is to
be called, debugger first calls \texttt{objectId()} on the
\texttt{foo} owner object. Then it stores this owner id, the stack
frame locations, and other state values in scope at the call point.
Whenever the execution reaches the reproduction point the debugger
looks at the history of \texttt{foo} changes and finds the last
\texttt{foo} change with the same object id as \texttt{bar} id at the
reproduction point. Figure ~\ref{fig:foo-changes1} shows the list of
property \texttt{foo} change events in a hypothetical
execution. \texttt{bar} id at the reproduction point is 1010, so the
last change of \texttt{bar.foo} is the fourth column.

\subsection{\textit{lastChange} on variable} 
In JavaScript, every frame has a scope chain and every available
variable in the frame comes from one of the scopes in the frame's
scope chain. Once a developer asks for the last change of variable
\texttt{foo} at the reproduction point, the debugger first determines the
variable's scope as follows: it iterates over the scopes in the scope
chain and the first scope which has a variable with the same name is
the variable's scope. There are five different scope types: local,
global, \texttt{with}, closure and \texttt{catch}. We explain these
cases in two groups.

\subsubsection{global and \texttt{with} scopes}
Global scope is the the most outer scope in the scope chain and it is
also referred to as the global object (the \texttt{window}
object in Web pages). This scope is a regular JavaScript object and therefore every
global variable is a property of global object. Similarly,
 \texttt{with} scopes are also regular objects. A \texttt{with}
scope is created by a \texttt{with()} block with an object as the
parameter. Every property of this parameter object is available inside the
block as a variable. \textit{lastChange} treats the case where variable's scope
is global or \texttt{with}, like \textit{lastChange} on an object
property.

@@

If the developer asks for the last change of variable \texttt{foo} at the 
reproduction point in a program, debugger calls 
\texttt{setVariableChangeH-} \texttt{ook()} with the variable's defining
block and name as parameters and re-executes the
program. Whenever \texttt{foo} changes and the callback function is to
be called, debugger first calls \texttt{scopeId()} on the
variable's scope. Then it stores this scope id, the stack
frame locations, and other state values in scope at the call point.
Whenever the execution reaches the reproduction point the debugger
looks at the history of \texttt{foo} changes and finds the last
\texttt{foo} change with the same scope id as the variable's scope id
at the reproduction point (Figure ~\ref{fig:foo-changes2}).

@@

\subsection{\textit{lastChange} on \textit{lastChange}}
The algorithm we explained so far needs an enhancement to support
cascades of \textit{lastChange}s. We explain first the issue
and then the solution. Consider the following points:

\begin{center}
\textit{
 point A : the reproduction point \\
 point B : lastChange(A, bar.x) \\
 point C : lastChange(B, baz.y) 
 }
 \end{center}
According to the algorithm explained in the previous section
the points B and C cannot be identified before reaching point
A. That is, as execution moves through point B we cannot know it will be the point of last change until
we reach A, and we cannot apply the simple algorithm to look for point C until we know B. 
 Therefore, all \texttt{x} change events for point B and all \texttt{y}
change events for point C are stored. Once the execution reaches point A,
and the exact \texttt{bar} id reveals, point B can be identified among
the change events in the history of \texttt{x} changes. The problem
arises in recognizing point C. To recognize point C, we need
\texttt{baz} id at point B, but B is a past program state, and
\texttt{baz} id has not been collected at this point.

To address this issue, we use Algorithm \ref{dependency-analysis}, to
create the list of additional data should be collected at a change
event. This algorithm does does a dependency analysis between all
\textit{lastChange} queries and computes the list of additional data
required to be collected at each change event. For example in the
mentioned case, \texttt{baz} id-if it is available- should be
collected at all points \texttt{x} changes. Having this additional
data, locating point C, using \texttt{baz} id at point B is
straightforward (Figure ~\ref{fig:lastchange-lastchange}).

@@

%---------------------------------------------------------------------------------------------------
\section{JavaScript implementation}
To verify the \textit{lastChange} algorithm we
implemented it in an extension to the Firebug
JavaScript debugger\cite{Firebug}. %\cite{firebug-version-1.6}. 
Firebug itself is an extension of the Firefox browser\cite{Firefox}. %\cite{firefox-version-3.6}
The Firefox JavaScript engine provides a JavaScript debugging interface \cite{JSD} and
\textit{Querypointer} is developed over this interface. Our prototype implementation defines our four primitive operations in JavaScript code and using techniques which are cumbersome and comparatively slow to execute.  A professionally useful debugger would implement these primitive operations in the Javasript engine.

@@

For our purposes, the \texttt{watch()} function only covers the case
of global object properties. At the beginning of execution, no object
excepting the global object is available.  For 
the our \textit{lastChange} prototype we created a version of
\texttt{setPropertyChangeHook()}.  The basic strategy is to get a reference to the object just
after its creation, then use \texttt{watch()} function to monitor
property changes in the object.  Setting a flag\cite{JSD} into the Firefox
JavaScript engine we can get file URL and line number for each object
creation (e.g., myFile.js, line 24). We set a hook on this line
and analyze the source code to determine which object was created.

The only data we have is the object creation location including the
file url and the line number and the goal is to get a reference to
this object. Although in the most regular cases we have only one
statement and one object creation in a line, there are cases that more
than one object might be created in a line. There is no way to
recognize the interesting object among these new objects. So instead
of one object, we monitor all new objects created in the line.

@@

There is another case for new objects and when they are passed as
arguments to other functions (Figure~\ref{fig:objectCreation}(c)). In these
cases instead of step-over we use step-in and we get the corresponding
argument as a new object.

Although this approach is successful in many ordinary cases, there are some 
cases that this approach doesn't work. For example in a case
where the new object is assigned to \texttt{a[++i]} where i is a variable
(Figure ~\ref{fig:objectCreation}(d)), \texttt{a[++i]} is kept in the list
of assignee statements. Obviously, evaluating this statement doesn't return
a reference to the new object. Our prototype implementation does not yet handle these kinds of unusual cases.

Throughout this section we implicitly assumed that the object will be created at 
the same location in the next execution. So, what does happens if it
is not true? Well, once the execution reaches the reproduction point, it
reveals that the object has been created in different location. This time,
debugger re-executes the program considering both locations as possible 
object creation locations.

\subsection{\texttt{scopeId()} operation}
The debugger sets a hook at the beginning of all code blocks needing
a scope id whenever the block is loaded. The scope id is kept as a variable with name 
\texttt{\_scopeId} in the scope. Whenever the hook hits,  meaning a new scope
is created, \texttt{\_scopeId} is set by calling JavaScript's dynamic compilation 
function \texttt{eval()}. For example, executing \texttt{eval( var \_scopeId = 10)} creates a
variable with name \texttt{\_scopeId} and value \texttt{10} in the
scope of the \texttt{eval()} call, which is our interesting scope. \texttt{scopeId()}
operation returns the value of \texttt{\_scopeId} in the scope.

\subsection{\texttt{setVariableChangeHook()} operation}
Variables defined in local, closure, and catch scopes are only changed in
their scope; that includes the defining scope and scopes nested inside of them.
For example in Figure ~\ref{fig:js-closure}, variable \texttt{var1} in line 11 can only be
changed in lines 10 to 23. Therefore, after locating the function in
which the variable is defined, it is enough to parse the code inside
the function block and set a hook on all lines where the variable is
assigned a new value. 
We also set a hook at the first line of the function which is
corresponding to the line variable is defined. In Figure
~\ref{fig:js-closure}, if the execution is paused at line 17 and the
last change of \texttt{var1} is queried, two hooks on lines 16 and 17
will be set, but if the execution is paused at line 20 , four hooks on
lines 11, 12, 14, 20 will be set. These two cases are different due
the fact that \texttt{var1} in \texttt{firstChild} is a local variable
but in \texttt{secondChild} is a closure variable.

@@

\section{Reproducible Non-deterministic Execution}

@@

Since \textit{lastChange} does not rely on determinism, some bugs that
confound breakpoint debuggers will be easier to find with
\textit{lastChange}.  Consider the example in 
Figure ~\ref{fig:counter-example}. It shows a function which processes an
array. The array contains numbers except one item which is
\texttt{undefined} and it causes a bug in line 30. There is a call to
\texttt{randomPermutation} function in line 11 which randomly
permutes the array item. So at every execution the
\texttt{undefined} item will be in a new place. Calling
\textit{lastChange} on \texttt{x} at the place bug happens, gives a
point which shows line 14. Although this point exists at every
re-execution but it can not be identified by a conditional breakpoint.

@@

We have presented the \textit{lastChange} algorithm and describe our
prototype implementation. Now we want to convince you that the
algorithm has practical value. Ultimately, practical value can only be
proven by developers in the field using the technique. This paper is
one step to convince implementers to put \textit{lastChange} in the
field.  Thus our goal here is a persuasive argument for potential implementors.

@@

\subsection{Developers need an operation like \protect\textit{lastChange} }

We have shown how \textit{lastChange} identifies the prior point in
program execution where a program state value changes. Is it something
the developer needs to do?  Since ultimately programs are just
transformation of state values, debugging is ultimately backtracking
to find defects in program state change.  When a developer halts a
program on a breakpoint they have three kinds of information: a model
of the program in their head, a call stack showing what part of the
program is halted, and the state at that point (a combination of the
debugger's view of the program internal state and the output of the
program up to this point of execution). If the model does not match
the call stack or if no value in the viewable program state aligns
with the developers model for this point in execution, the developer
will seek another view by changing the breakpoint or the input
data. If a value appears incorrect, they may have a flash of insight
and know the defect. Otherwise they need to figure out what operation
causes the incorrect value. Here \textit{lastChange} takes over and addresses a key
part of the debugging process.

\subsection{Developers can learn to use \protect\texttt{lastChange}}

@@

\subsection{Practical implementations are feasible}

@@

\subsection{In most cases \protect\texttt{lastChange} will be much faster than current alternatives}

@@

We could also compare to solutions based on logging or tracing. Manual
logging has very high overhead: the developer must add code, debug
that added code, then analyze the log. (To be fair, the log can become
a permanent debugging aid.) Automatic logging as we discussion in
\ref{sec:relatedWork} causes about one or two orders of magnitude slow down.

\subsection{The worst cases are not more common or more painful than alternatives}

@@

\section{Related Work}
\label{sec:relatedWork}

@@

\section{Conclusion}
\textit{lastChange} provides critical information for debugging programs: the location and state at the point where a questionable value was assigned. It builds upon existing technology and developer experience making it a practical solution for implementers. Our prototype demonstrates the feasibility of \textit{lastChange} and its user interface hints at the potential this approach can have in organizing the debugging experience.
